Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-05-17 Docs/FDAgroup.Rmd
calculated AUC for NHANES data

# df = content
# df_train = subset(content, include == 1)
# df_test = subset(content, include == 0)
```

@@

<!-- ## Simulated child growth data -->

<!-- - Predict length-for-age z-score (LAZ), observed with noise -->
<!--   - Calculated using the age- and sex-specific WHO standard references -->

<!-- ```{r} -->
<!-- ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+ -->
<!--   geom_line(show.legend = F)+ -->
<!--   labs(title = "Length-for-Age z-score tracks for the 10 subjects", -->
<!--        x = "time", y ="LAZ") -->
<!-- ```  -->

<!-- ## FPCA on observed LAZ -->

<!-- ```{r} -->
<!-- tnew = sort(unique(df_train$argvals)) -->
<!-- df_laz <- df_train %>% select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") -->
<!-- fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15, -->
<!--                         argvals.new = tnew, calculate.scores = TRUE) -->

<!-- # mean function -->

<!-- par(mfrow = c(2, 2)) -->
<!-- plot(tnew, fpca_laz$mu.new, type = "l",  -->
<!--      xlab = "t", ylab = "f0(t)") -->
<!-- plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1") -->
<!-- plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2") -->
<!-- plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3") -->

<!-- ## Prediction of new partially observed sample -->

<!-- ```{r, pred_func} -->
<!-- # write a function for prediction up to t -->
<!-- # df_test: datafram with columns argvals, subj, y, observation up to tm -->

<!-- DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){ -->
<!--   Ntest = length(unique(df_test$subj)) -->
<!--   # prediction -->
<!--   pred1 <- predict(fpca_fit, newdata = df_test) -->
<!--   # Eigen functions -->
<!--   Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ] -->
<!--   Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores) -->
<!--   # every column is a new function -->
<!--   Yhat <- data.frame(Yhat) -->
<!--   colnames(Yhat) <- pred1$rand_eff$subj -->
<!--   df_pred <- Yhat %>%  -->
<!--     mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>%  -->
<!--     pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>% -->
<!--     mutate(type = "pred") -->

<!--   return(df_pred) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r t0_3} -->
<!-- df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") %>%  -->
<!--   filter(argvals <= 0.3)) -->

<!-- # figure  -->
<!-- bind_rows(df_test %>%  -->
<!--             select(argvals, subj, Y)%>%  -->
<!--             rename("y" = "Y") %>%  -->
<!--             mutate(type = "true", subj = as.factor(subj)),  -->
<!--           df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>%  -->
<!--   filter(subj == 152) %>%  -->
<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
<!--   geom_line()+ -->
<!--   facet_wrap(~subj)+ -->
<!--   labs(title = "Prediction with observation up to t=0.3",  -->
<!--        x = "time", y = "") -->
<!-- ``` -->

<!-- ## Prediction of new partially observed sample -->

<!-- ```{r} -->
<!-- df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") %>%  -->
<!--   filter(argvals <= 0.5), tm = 0.5) -->

<!-- # figure  -->

<!-- bind_rows(df_test %>%  -->
<!--             select(argvals, subj, Y)%>%  -->
<!--             rename("y" = "Y") %>%  -->
<!--             mutate(type = "true", subj = as.factor(subj)),  -->
<!--           df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>%  -->
<!--   filter(subj %in% 151:156) %>%  -->
<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
<!--   geom_line()+ -->
<!--   facet_wrap(~subj)+ -->
<!--   labs(title = "Prediction with observation up to t=0.5",  -->
<!--        x = "time", y = "") -->

<!-- ## Prediction of new partially observed sample -->

<!-- ```{r} -->
<!-- df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") %>%  -->
<!--   filter(argvals <= 0.8), tm = 0.8) -->

<!-- # figure  -->
<!-- bind_rows(df_test %>%  -->
<!--             select(argvals, subj, Y)%>%  -->
<!--             rename("y" = "Y") %>%  -->
<!--             mutate(type = "true", subj = as.factor(subj)),  -->
<!--           df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>%  -->
<!--   filter(subj %in% 151:156) %>%  -->
<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
<!--   geom_line()+ -->
<!--   facet_wrap(~subj)+ -->
<!--   labs(title = "Prediction with observation up to t=0.8",  -->
<!--        x = "time", y = "") -->

>>>>

df = content
df_train = subset(content, include == 1)
df_test = subset(content, include == 0)
```

@@

## Simulated child growth data

- Predict length-for-age z-score (LAZ), observed with noise
  - Calculated using the age- and sex-specific WHO standard references

```{r}
ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+
  geom_line(show.legend = F)+
  labs(title = "Length-for-Age z-score tracks for the 10 subjects",
       x = "time", y ="LAZ")
```

## FPCA on observed LAZ

```{r}
tnew = sort(unique(df_train$argvals))
df_laz <- df_train %>% select(argvals, subj, Y) %>%
  rename("y" = "Y")
fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15,
                        argvals.new = tnew, calculate.scores = TRUE)

# mean function

par(mfrow = c(2, 2))
plot(tnew, fpca_laz$mu.new, type = "l",
     xlab = "t", ylab = "f0(t)")
plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")

```

## Prediction of new partially observed sample

```{r, pred_func}
# write a function for prediction up to t
# df_test: datafram with columns argvals, subj, y, observation up to tm

DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){
  Ntest = length(unique(df_test$subj))
  # prediction
  pred1 <- predict(fpca_fit, newdata = df_test)
  # Eigen functions
  Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ]
  Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores)
  # every column is a new function
  Yhat <- data.frame(Yhat)
  colnames(Yhat) <- pred1$rand_eff$subj
  df_pred <- Yhat %>%
    mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>%
    pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>%
    mutate(type = "pred")

  return(df_pred)
}
```

```{r t0_3}
df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>%
  filter(argvals <= 0.3))

# figure
bind_rows(df_test %>%
            select(argvals, subj, Y)%>%
            rename("y" = "Y") %>%
            mutate(type = "true", subj = as.factor(subj)),
          df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>%
  filter(subj == 152) %>%
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.3",
       x = "time", y = "")
```

## Prediction of new partially observed sample

```{r}
df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>%
  filter(argvals <= 0.5), tm = 0.5)

# figure

bind_rows(df_test %>%
            select(argvals, subj, Y)%>%
            rename("y" = "Y") %>%
            mutate(type = "true", subj = as.factor(subj)),
          df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>%
  filter(subj %in% 151:156) %>%
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.5",
       x = "time", y = "")

```

## Prediction of new partially observed sample

```{r}
df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>%
  filter(argvals <= 0.8), tm = 0.8)

# figure
bind_rows(df_test %>%
            select(argvals, subj, Y)%>%
            rename("y" = "Y") %>%
            mutate(type = "true", subj = as.factor(subj)),
          df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>%
  filter(subj %in% 151:156) %>%
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.8",
       x = "time", y = "Length for age z-score")

```

```{r}
bind_rows(df_test %>%
            select(argvals, subj, Y)%>%
            rename("y" = "Y") %>%
            mutate(type = "true", subj = as.factor(subj)),
          df_test_t0.3 %>% mutate(subj = as.factor(subj),
                                  type = "Up to t=0.3"),
          df_test_t0.5 %>% mutate(subj = as.factor(subj),
                                  type = "Up to t=0.5"),
          df_test_t0.8 %>% mutate(subj = as.factor(subj),
                                  type = "Up to t=0.8")) %>%
  filter(subj == 152) %>%
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  labs(title = "Dynamic prediction of child growth data",
       x = "Time", y = "Length for age z-score")

ggsave(here("Images/dynpred_exp.png"), width = 8, height = 6, bg = "white")

```

### 2023-05-17 Docs/Manuscript_edit.Rmd
calculated AUC for NHANES data

# fGFPCA
fig1a <- exp_df_sim1 %>%
  left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i, col = "True"))+
  geom_line(aes(x=bin, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))

# GLMMadaptive
fig1b <- exp_df_sim1_ref %>%
  filter(sind_inx %in% mid) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=sind_inx, y=eta_i, col = "True"))+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))

@@

ave_ise_mat <- apply(ise_mat, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat) <- c("200", "400", "600", "800")
rownames(ave_ise_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

@@

ave_ise_mat_ref <- apply(ise_mat_ref, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat_ref) <- c("200", "400", "600", "800")
rownames(ave_ise_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

@@

ave_auc_mat <- apply(auc_mat, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat) <- c("200", "400", "600", "800")
rownames(ave_auc_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

@@

ave_auc_mat_ref <- apply(auc_mat_ref, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat_ref) <- c("200", "400", "600", "800")
rownames(ave_auc_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

@@

```{r, Figure_appl, fig.cap="Predicted tracks of four randomly selected subjects from the daily weigh-in dataset. The grey points are observed binary outcomes. The dashed lines indicate predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
rand_id2 <- sample(N2, 4)
fig2a <- Y_df2 %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_100, pred_200, pred_300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Bin", 
       colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))

fig2b <- df_est_latent_ref %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t100, pred_t200, pred_t300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_t100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Bin", 
       colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))

>>>>

# fGFPCA
fig1a <- exp_df_sim1 %>%
  left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i, col = "True"))+
  geom_line(aes(x=bin, y = pred_t200, col = "Up to 0.2"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "Up to 0.4"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "Up to 0.6"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col= "Up to 0.8"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Time", colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000), labels = c(0, 0.2, 0.4, 0.6, 0.8, 1.0))

# GLMMadaptive
fig1b <- exp_df_sim1_ref %>%
  filter(sind_inx %in% mid) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=sind_inx, y=eta_i, col = "True"))+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 0.2"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t400, col = "Up to 0.4"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t600, col = "Up to 0.6"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t800, col= "Up to 0.8"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Time", colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000),
                     labels = c(0, 0.2, 0.4, 0.6, 0.8, 1.0))

@@

ave_ise_mat <- apply(ise_mat, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat) <- c("0.2", "0.4", "0.6", "0.8")
rownames(ave_ise_mat) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")

@@

ave_ise_mat_ref <- apply(ise_mat_ref, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat_ref) <- c("0.2", "0.4", "0.6", "0.8")
rownames(ave_ise_mat_ref) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")

@@

ave_auc_mat <- apply(auc_mat, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat) <- c("0.2", "0.4", "0.6", "0.8")
rownames(ave_auc_mat) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")

@@

ave_auc_mat_ref <- apply(auc_mat_ref, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat_ref) <- c("0.2", "0.4", "0.6", "0.8")
rownames(ave_auc_mat_ref) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")

@@

```{r, Figure_appl, fig.cap="Predicted tracks of four randomly selected subjects from the daily weigh-in dataset. The grey points are observed binary outcomes. The dashed lines indicate predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
rand_id2 <- sample(N2, 4)
fig2a <- Y_df2 %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_100, pred_200, pred_300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Time", 
       colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))

fig2b <- df_est_latent_ref %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t100, pred_t200, pred_t300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_t100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Time", 
       colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))
